Using CopyKAT to idenfity tumor cells in Ewing sarcoma samples

### 2024-05-21 analyses/cell-type-ewings/template_notebooks/03-copykat.Rmd
update template notebook to take in new params

---
title: "Using CopyKAT to idenfity tumor cells in Ewing sarcoma samples"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
params:
  sample_id: SCPCS000490
  library_id: SCPCL000822
---

@@

`CopyKAT` was run using the `run-copykat.R` script and specifying endothelial cells from both `SingleR` and `CellAssign` as the normal reference. 
This produced results from `CopyKAT` both with and without a normal reference. 
These results are read into this notebook and used to:

@@

# tumor/normal classifications 
classifications_filename <- glue::glue("{params$library_id}_tumor_normal_classifications.tsv")
classifications_file <- file.path(module_base, "results", "marker_gene_analysis", classifications_filename)

# copykat results 
results_dir <- file.path(module_base, "results", "copykat", params$library_id)

# predictions files 
predictions_file <- glue::glue("{params$library_id}_copykat_prediction.txt")

predictions_paths <- c(
  no_ref = file.path(results_dir, "no_normal", predictions_file),
  with_ref = file.path(results_dir, "endothelial_normal", predictions_file)
)

# full results gene by cell 
full_ck_result_file <- glue::glue("{params$library_id}_copykat_CNA_results.txt")

full_ck_result_paths <- c(
  no_ref = file.path(results_dir, "no_normal", full_ck_result_file),
  with_ref = file.path(results_dir, "endothelial_normal", full_ck_result_file)
)

```

@@

# read in tumor normal classifications 
manual_classifications_df <- readr::read_tsv(classifications_file)

# read in ck predictions from both reference types (no_normal and with_normal)
ck_results_df <- predictions_paths |> 
  purrr::map(readr::read_tsv) |> 
  dplyr::bind_rows(.id = "reference_used")

# read in full gene by cell copy number detection results 
full_ck_results_df <- full_ck_result_paths |> 
  purrr::map(readr::read_tsv) |> 
  dplyr::bind_rows(.id = "reference_used")
```

@@

![](`r file.path(results_dir, "no_normal", "SCPCL000822_copykat_heatmap.jpeg")`)

@@

![](`r file.path(results_dir, "endothelial_normal", "SCPCL000822_copykat_heatmap.jpeg")`)

Both of these heatmaps appear to show some increase in observed CNVs in aneuploid vs. diploid which is to be expected.

@@

If we look at the UMAPs shown in `01-marker-gene-tumor-classification.Rmd`, the cells that are labeled aneuploid here are consistent with where we saw marker gene expession for tumor cells. 
It also looks like there is not a visible difference between using a reference of normal cells and not using a reference.

@@

It looks like the largest increases are in chr8 and chr21. 
Chr8 is one of the more common CNA's in Ewing's so this is a good sign.

Chr12 may show some increase, but Chr1 and Chr16 don't show any real differences.

@@

Looking at these results, we see that cells classified as tumor cells using marker genes are mostly overlapping with cells that are aneuploid in CopyKAT, which is what we expect! 
We also see that the Kappa values are close to 1, indicating almost perfect agreement.

It also looks like there isn't a difference between using the reference and not using the reference. 
The reference Kappa is .1 higher, but I don't think that's enough to argue that you need to use a reference. 
The one thing to note is that using a reference cut the run time for `CopyKAT` in half, so that is one benefit.

We can also calculate the Jaccard similarity index to visualize the amount of cells that have overlapping annotations. 
Given the confusion matrix, I would expect to see good agreement between marker gene and `CopyKAT` annotations.

@@

As expected, it looks like cells classified as tumor cells using marker gene annotation are classified as `aneuploid` with `CopyKAT`.

@@

We see that aneuploid cells are mostly muscle cells in both `CellAssign` and `SingleR`. 
In `SingleR` there are some other cell types that were assigned that have a few anueploid cells (chondrocytes, neurons, and fibroblasts).

For diploid cells, we see the most overlap with fibroblasts, endothelial cells, and macrophages. 
These are consistent with the cell types that we see normal cells line up with in `01-marker-gene-tumor-classification.Rmd`.

## Conclusions

- `CopyKAT` seems to do a good job in this Ewing sarcoma sample of identifying aneuploid and diploid cells. 
- There is some small evidence of copy number gains in Chr8 and Chr12 which is consistent with findings in Ewing sarcoma patients. 
- The annotations from `CopyKAT` mostly line up with annotations obtained using marker gene annotation. 
- Using a reference of normal cells doesn't change the overall results but does reduce the run time for `CopyKAT`.

>>>>

---
title: "Using CopyKAT to idenfity tumor cells in Ewing sarcoma samples"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
params:
  sample_id: SCPCS000490
  library_id: SCPCL000822
  marker_gene_classification: "../results/annotate_tumor_cells_output/SCPCS000490/SCPCL000822_tumor-normal-classifications.tsv"
  no_ref_copykat_results: "../results/annotate_tumor_cells_output/SCPCS000490/no_reference"
  with_ref_copykat_results: "../results/annotate_tumor_cells_output/SCPCS000490/with_reference"
---

@@

`CopyKAT` was run using the `run-copykat.R` script with and without a normal reference. 
These results are read into this notebook and used to:

@@

obj_file <- glue::glue("{params$library_id}_final-copykat.rds")
png_file <- glue::glue("{params$libary_id}_heatmap-copykat.png")

copykat_objs <- c(
  no_ref = file.path(params$no_ref_copykat_results, obj_file),
  with_ref = file.path(params$with_ref_coypkat_results, obj_file)
) |> 
  purrr::map(readr::read_rds)

png_list <- c(
  no_ref = file.path(params$no_ref_copykat_results, png_file),
  with_ref = file.path(params$with_ref_copykat_results, png_file)
)
```

@@

# read in tumor normal classifications 
manual_classifications_df <- readr::read_tsv(params$marker_gene_classification)

# read in ck predictions from both reference types (no_normal and with_normal)
ck_results_df <- copykat_objs |> 
  purrr::map(\(obj){
    obj$prediction |> 
      as.data.frame()
  }) |> 
  dplyr::bind_rows(.id = "reference_used")

# read in full gene by cell copy number detection results 
full_ck_results_df <- copykat_objs |> 
  purrr::map(\(obj){
    obj$CNAmat |> 
      as.data.frame()
  }) |> 
  dplyr::bind_rows(.id = "reference_used")
```

@@

![](`r png_list["no_ref"]`)

@@

![](`r png_list["with_ref"]`)9

@@

@@

@@

We can also calculate the Jaccard similarity index to visualize the amount of cells that have overlapping annotations.

@@

@@
